How topos theory can help algebra and geometry

### 2018-07-10 slides-leipzig2018.tex
Beginning of notes for the Leipzig talk

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

      \footnotesize
      ``Let~$G$ be a group.''
    \end{column}
  \end{columns}
\end{frame}

@@

    \begin{scope}[visible on=<1->]
      \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1) (more-set1)}{}
    \end{scope}
    \begin{scope}[visible on=<2->]
      \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1) (more-eff1)}{tape}
    \end{scope}
    \begin{scope}[visible on=<2->]
      \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
      \def\R{8pt}
      \begin{pgfonlayer}{background}
      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple, visible on=<2->]
        ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
      \end{pgfonlayer}
    \end{scope}
    \begin{scope}[visible on=<3->]
      \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}{}
    \end{scope}
  \end{tikzpicture}
\end{frame}

>>>>

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\footnotesize\justifying#2\par}}}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

      \footnotesize
      ``Let~$G$ be a group.''
    \end{column}
  \end{columns}

  \jnote{1}{Toposes were invented in the 1960s by Grothendieck in order to
  solve concrete problems in algebraic geometry. Their raison d'être is the
  following. In algebraic geometry, we want (and sometimes have to) work over
  fields other than~$\RR$ and~$\CC$ and even over arbitrary commutative rings.
  For the geometric objects in such settings, the \emph{schemes}, the Euclidean
  topology is not available; we have to make do with the \emph{Zariski
  topology}. However, important tools as singular cohomology don't work well
  with this topology.

  The problem was solved by inventing \emph{étale topology} as an enhancement
  of the Zariski topology. However, contrary to its name, the étale topology
  isn't a true topology in the usual sense. Putting the étale topology on a
  scheme doesn't yield a refined topological space, but a \emph{topos}.

  Toposes generalize topological spaces in two ways: Firstly, the ``open sets''
  of toposes don't actually have to be sets of points; they can be more general
  kinds of objects such as coverings. Toposes can even have no points at all
  and still be nontrivial (this is for instance the case for the \emph{topos of
  random sequences}). Secondly, while classically a given open
  subset is either contained in a further open subset or not, the opens of
  toposes can be contained in further opens in many different ways.

  Recently, toposes are being used to help the mysterious field with
  one element come into being.}

  \jnote{2}{Since the 1960s, many more aspects of toposes were discovered.
  A popular reference on toposes starts with a list of 13 ways of viewing
  toposes.

  In this talk, we'll focus on the view of toposes as \emph{alternate
  mathematical universes}. We can do mathematics inside these alternate
  universes just as well as we can do mathematics inside the \emph{standard
  universe} (which is represented by a particular topos called~``$\Set$'').
  Each topos contains own versions of all the familiar mathematical objects --
  numbers, functions, manifolds -- but the properties they enjoy differ
  slightly from topos to topos.

  The definition of what a topos is, displayed in the lower left of the slide,
  has two problems. Firstly, it's only useful if one knows the relevant
  category-theoretic jargon. Secondly, a topos has lots of further vital
  structure, which is crucial for a rounded understanding. However, the
  displayed definition is trimmed for minimality and therefore doesn't mention
  it. A more comprehensive definition is: A \emph{topos} is a locally cartesian
  closed, finitely complete and cocomplete Heyting category which is exact,
  extensive and has a subobject classifier. We won't need either definition in
  this talk.}

  \jnote{3}{Toposes can also be thought as reifying the ``semantic essence'' of
  mathematical theories (the theory of groups, the theory of rings, the theory
  of intervals, \ldots). Given any such theory~$T$, there is a so-called
  \emph{classifying topos}~$\Set[T]$. Its points are precisely the set-based
  models of~$T$ (the groups, the rings, the intervals, \ldots), and it contains
  a \emph{generic model} which has exactly those properties which all models
  have. (This generic model is what mathematicians implicitly refer to when
  they say ``Let~$G$ be a group.''.)

  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
  toposes even when they are not syntactially related in any way. This
  observation is the starting point of Olivia Caramello's \emph{bridge
  technique}, a grand research program with applications in many different
  fields.

  \[ \xymatrix{
    & \Set[T] \simeq \Set[T'] \ar@{--}@/^1pc/[rd] \ar@{--}@/_1pc/[ld] \\
    T && T'
  } \]}
\end{frame}

@@

    \begin{scope}[visible on=<1->]
      \drawbox{set1}{$\mathrm{Set}$}{(objs-set1) (prop-set1) (more-set1)}{}
    \end{scope}
    \begin{scope}[visible on=<2->]
      \drawbox{eff1}{Ef{}f}{(objs-eff1) (prop-eff1) (more-eff1)}{tape}
    \end{scope}
    \begin{scope}[visible on=<2->]
      \drawbox{sh1}{$\mathrm{Sh}\, X$}{(objs-sh1) (prop-sh1) (more-sh1)}{draw=none}
      \def\R{8pt}
      \begin{pgfonlayer}{background}
      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple, visible on=<2->]
        ($(sh1.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(sh1.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(sh1.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(sh1.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
      \end{pgfonlayer}
    \end{scope}
    \begin{scope}[visible on=<3->]
      \drawbox{set0}{$\mathrm{Set}$}{(objs-set0) (set1) (eff1) (sh1)}{}
    \end{scope}
  \end{tikzpicture}

  \jnote{1}{The topos~$\Set$ is the standard topos. This topos is where
  ordinary mathematics takes place.}

  \jnote{2}{Besides~$\Set$, there is a proper class worth of further toposes.
  We'll get to know some of these better during the remaining course of the
  talk.

  For any topological space~$X$, there is a topos~$\Sh(X)$ of \emph{sheaves
  over~$X$}. They are useful in analysis for internalizing
  parameter-dependence. Apart from pathological cases like~$X$
  being discrete, the intermediate value theorem in the form
  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then there is a number~$x \in \RR$ such
  that~$f(x) = 0$.\end{indentblock}
  fails in these toposes -- for very meaningful reasons discussed below.
  The following, classically equivalent, version does hold:
  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then, for every~$\varepsilon > 0$, there is
  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.\end{indentblock}

  The \emph{effective topos}~$\Eff$ is a computer scientist's dream come true:
  In it, any function~$\NN \to \NN$ is computable by a Turing machine. The
  effective topos and its close cousins can be used to study the differences
  between the many models of computation, particularly those differences which
  are only visible at higher types.}
\end{frame}
